---
title: "Contours and hydraulic-gradient arrows"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contours and hydraulic-gradient arrows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white")
```

```{r surface}
library(potentiomap)
data("synthetic_wells")
points <- ps_make_points(
  synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916"
)
surface <- ps_interpolate(points, methods = "IDW", grid_res = 300)$IDW
```

The default contour result is a line `SpatVector`. A structured result adds an
inventory of every requested level.

```{r contours}
contours <- ps_contours(
  surface, levels = c(166, 168, 170), return = "result"
)
contours$manifest
```

Explicit levels outside the finite surface range remain in the manifest and
produce a classed warning. Open lines are not silently closed or converted to
polygons.

Individual contour lines can be divided where local support changes.

```{r contour-support}
support <- ps_prediction_support(points, surface = surface)
classified <- suppressWarnings(ps_contour_support(
  contours$contours, support = support,
  supported_distance = 500, approximate_distance = 1200,
  require_inside_hull = TRUE
))
classified$summary
```

```{r contour-support-map, fig.width=7, fig.height=4, fig.alt="Synthetic potentiometric contours divided into solid supported sections and dashed approximate sections, with groundwater observation wells shown as points."}
plot(classified, show_unsupported = FALSE, legend_position = "topright")
terra::plot(points, add = TRUE, pch = 20)
```

The `line_type` field suggests solid, dashed, or dotted symbology while keeping
the support class and reason as GIS attributes. The classes use thresholds
chosen by the user, not confidence intervals. Proximity to wells does not prove
that a contour is correct, distance does not prove that it is wrong, and every
section remains an interpolation from the modeled surface. Coarse support cells
can shift the apparent solid-to-dashed transition.

Hydraulic-gradient arrows point along the local negative modeled-head gradient.
The following deliberately uses short, sparse display symbols.

```{r arrows}
flow <- ps_flow_arrows(
  surface, scale = 25, res_factor = 10, endpoint_action = "shorten"
)
flow$validation_summary
head(flow$validation)
```

`endpoint_action = "flag"` preserves failing lines and reports them.
`"shorten"` repeatedly halves their length along the original direction.
`"drop"` removes failures. `"none"` retains the version 0.1.0 unvalidated
geometry for comparison. No policy reverses or bends an arrow.

Arrow length is a cartographic convention based on gradient, cell size, and the
chosen scale. The line is not a velocity, travel time, particle path, or traced
groundwater path. Endpoint checks only evaluate a straight symbol against the
supplied raster; they do not prove that the interpolated surface is physically
correct.
