---
title: "Units, vertical references, and grouped groundwater observations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Units, vertical references, and grouped groundwater observations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white")
```

Groundwater elevation calculated from depth requires an explicit sign and
measurement reference. With positive depth below land surface:

`groundwater elevation = land-surface elevation - depth to water`.

```{r units}
library(potentiomap)
data("synthetic_wells")
points <- ps_potentiometric_points(
  synthetic_wells,
  x = "x", y = "y", depth_col = "depth_to_water",
  surface_col = "surface_elevation", name_col = "well_id",
  crs = "EPSG:26916", depth_unit = "m", surface_unit = "m",
  output_unit = "ft", vertical_datum = "synthetic example datum",
  surface_reference = "land_surface", depth_sign = "positive_down",
  metadata_mode = "strict"
)
ps_metadata(points)
knitr::kable(data.frame(
  well = head(terra::values(points)$Name),
  groundwater_elevation_ft = head(terra::values(points)$Z)
))
```

The supported length units are metres and international feet, using exactly
`1 ft = 0.3048 m`. A horizontal CRS does not supply a vertical datum.
potentiomap records the stated datum but does not transform between vertical
datums, infer datum compatibility, or guess measuring-point offsets. R-level
metadata may not survive every GIS vector format, so preserve a sidecar manifest
when those details must accompany exports.

Separate events and water-bearing units before fitting surfaces.

```{r grouped}
grouped_data <- transform(
  synthetic_wells,
  event = rep(c("spring", "autumn"), each = 16),
  unit = rep(c("upper", "lower"), times = 16)
)
grouped <- ps_interpolate_grouped(
  grouped_data, c("event", "unit"), value = "gw_elevation",
  name_col = "well_id", crs = "EPSG:26916",
  methods = "IDW", grid_res = 400
)
knitr::kable(grouped$manifest)
```

Each row is assigned to exactly one group. A shared template standardizes output
geometry but does not share observations or fitted information. Empty and failed
groups remain in the manifest. A simple progress callback can report group
completion without adding another package dependency.
