---
title: "Citing potentiomap"
description: "CRAN availability, DOI, package version, license, citation, and BibTeX."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Citing potentiomap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The website documents **potentiomap 0.2.0**. Version 0.2.0 has been submitted
to CRAN; use the CRAN page and the installed package citation to confirm the
currently available release.

```{r release-metadata, echo=FALSE}
knitr::kable(data.frame(
  item = c("Website version", "CRAN package", "Permanent DOI", "License"),
  value = c(
    "0.2.0",
    "potentiomap",
    "10.32614/CRAN.package.potentiomap",
    "GPL-3"
  )
))
```

- CRAN: <https://CRAN.R-project.org/package=potentiomap>
- DOI: <https://doi.org/10.32614/CRAN.package.potentiomap>
- Source: <https://github.com/el-cordero/potentiomap>
- License: GPL-3

## Recommended citation

Always prefer the result returned by the installed version:

```{r citation-output}
citation("potentiomap")
```

The printed citation includes BibTeX output when used through
`toBibtex(citation("potentiomap"))`. Authorship and availability information
comes from the installed package metadata; no affiliation or biography is
added by this website.
