---
title: "Complete synthetic aquifer example"
description: "A complete fixed-seed 0.2.0 workflow from observation QA through validation, uncertainty, network review, profiles, and export."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Complete synthetic aquifer example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.bg = "white",
  fig.width = 9,
  fig.height = 6,
  fig.align = "center",
  dpi = 144,
  warning = TRUE,
  message = TRUE
)
library(potentiomap)
library(terra)
set.seed(20260717)

head_cols <- hcl.colors(64, "RdYlBu", rev = TRUE)
diff_cols <- hcl.colors(64, "Blue-Red 3", rev = TRUE)

draw_flow <- function(x, col = "#17252d", length = 0.045) {
  base <- crds(ps_arrow_vertices(x, "first"))
  tip <- crds(ps_arrow_vertices(x, "last"))
  length_xy <- sqrt((base[, 1] - tip[, 1])^2 + (base[, 2] - tip[, 2])^2)
  keep <- is.finite(base[, 1]) & is.finite(base[, 2]) &
    is.finite(tip[, 1]) & is.finite(tip[, 2]) & length_xy > 1
  arrows(base[keep, 1], base[keep, 2], tip[keep, 1], tip[keep, 2],
         length = length, angle = 24, col = col)
}
```

This article is a reproducible tour of the 0.2.0 analysis objects. Every input
comes from the installed package, every stochastic operation uses an explicit
seed, and export examples write only to a temporary directory. The example is
deliberately broad: a real project should run only analyses that answer its
defined hydrogeologic and monitoring questions.

<div class="pm-science-note">
This is a synthetic demonstration, not a site conceptual model. Interpolation does not create hydrogeologic knowledge outside the observations, and potentiomap is not a groundwater-flow or contaminant-transport simulator.
</div>

## 1. Define the analysis and load installed fixtures

We use a projected CRS whose horizontal coordinates are metres and retain the
fixture's head units as metres. The two monitoring events use the same vertical
datum and water-bearing-unit label. Those compatibilities must be verified,
not assumed, for field data.

```{r load-data}
data("synthetic_events", package = "potentiomap")
data("synthetic_wells", package = "potentiomap")
data("synthetic_nested_wells", package = "potentiomap")
data("synthetic_regions", package = "potentiomap")
data("synthetic_candidate_sites", package = "potentiomap")
data("synthetic_anisotropic_points", package = "potentiomap")
data("synthetic_transect", package = "potentiomap")

spring <- subset(synthetic_events, event == "spring")
autumn <- subset(synthetic_events, event == "autumn")
attr(spring, "crs") <- "EPSG:26916"

data.frame(
  fixture = c("spring event", "autumn event", "survey wells", "nested intervals"),
  records = c(nrow(spring), nrow(autumn), nrow(synthetic_wells),
              nrow(synthetic_nested_wells))
)
```

```{r observation-map, fig.alt="Synthetic spring monitoring wells colored from blue at lower head to red at higher head across a projected aquifer study area."}
spring_points <- ps_make_points(
  spring, "x", "y", "head", "well_id", "EPSG:26916"
)
plot(spring_points, "Z", col = head_cols, pch = 21, cex = 1.1,
     main = "Synthetic spring monitoring event")
```

## 2. Check observations before fitting a surface

`ps_check_observations()` reports issues without quietly redefining the input.
The check covers numeric coordinates and heads, identifiers, dates, units, and
vertical-datum fields. A clean synthetic result is not evidence that field data
would pass the same review.

```{r qa}
qa <- ps_check_observations(
  spring,
  x = "x", y = "y", value = "head", id = "well_id",
  datetime = "datetime", unit = "unit", vertical_datum = "vertical_datum"
)
qa$summary
qa$issues
```

Proceed only after resolving errors and documenting accepted review findings.
Unit conversion, datum reconciliation, casing-reference corrections, and
screen-interval interpretation belong upstream of interpolation.

## 3. Select a coherent monitoring event

The event selector records the target time, window, actual selected span,
ties, and excluded wells. Here the requested three-hour window around noon
captures one reading per well.

```{r event-selection}
event <- ps_select_event(
  spring,
  id = "well_id", datetime = "datetime",
  center = as.POSIXct("2025-03-15 12:00", tz = "UTC"),
  window = 3 * 60 * 60,
  rule = "nearest"
)
event$summary
head(event$selected[, c("well_id", "datetime", "head", "quality")])
```

The event window is a design choice. A short span does not guarantee hydraulic
synchroneity where pumping, tides, recharge, barometric response, or aquifer
storage cause meaningful within-event change.

## 4. Keep screened intervals and water-bearing units explicit

Nested intervals should not be pooled merely because they share coordinates.
The rules below classify intervals by vertical overlap with two declared
ranges. Ambiguous and unclassified records remain inspectable.

```{r screen-groups}
screen_rules <- data.frame(
  unit = c("upper", "lower"),
  top = c(150, 130),
  bottom = c(130, 110)
)

screen_groups <- ps_screen_groups(
  synthetic_nested_wells,
  mode = "rules",
  screen_top = "screen_top",
  screen_bottom = "screen_bottom",
  rules = screen_rules,
  overlap_required = 0.5
)
screen_groups$summary
head(screen_groups$assigned[, c("nest_id", "interval", "unit")])
```

Grouping is a transparent classification, not proof that an interval belongs
to a hydraulically connected unit. That interpretation requires the site
conceptual model and supporting stratigraphic evidence.

## 5. Fit alternative surfaces and retain diagnostics

We use the first 18 survey wells so the example runs quickly. The returned
`potentiomap_result` keeps the prediction template, fits, conditions,
diagnostics, method parameters, and support products alongside the surfaces.
Messages and numerical fit warnings are printed deliberately because they are
part of the audit trail.

```{r interpolate}
points <- ps_make_points(
  synthetic_wells[1:18, ],
  x = "x", y = "y", value = "gw_elevation",
  name_col = "well_id", crs = "EPSG:26916"
)

fit <- ps_interpolate(
  points,
  methods = c("IDW", "TPS", "OK"),
  grid_res = 300,
  support = TRUE,
  return = "result"
)

fit$method_parameters
fit$conditions
```

```{r method-surfaces, fig.width=10, fig.height=4, fig.alt="Side-by-side IDW, thin-plate spline, and ordinary kriging potentiometric surfaces using the same synthetic wells, grid, and head color range."}
common_range <- range(unlist(lapply(fit$surfaces, minmax)), finite = TRUE)
par(mfrow = c(1, 3), mar = c(2.6, 2.6, 3, 1))
for (method in c("IDW", "TPS", "OK")) {
  plot(fit$surfaces[[method]], col = head_cols, range = common_range,
       main = method, axes = FALSE)
  plot(points, add = TRUE, pch = 21, bg = "white", cex = 0.6)
}
```

The three maps are alternative modeled representations, not competing truths.
Differences are most consequential where observations are sparse or where a
method's trend and covariance assumptions control extrapolation.

## 6. Validate the declared prediction task and tune inside it

This three-fold example asks how well a method predicts held-out wells drawn
from this network. It does not measure continuous area-wide accuracy. Spatial
blocks, leave-cluster-out folds, user folds, or independent data can be more
appropriate for other tasks.

```{r validation}
validation <- ps_validate(
  points,
  methods = c("IDW", "TPS"),
  design = "kfold",
  folds = 3,
  prediction_mode = "direct",
  seed = 21
)
comparison <- ps_compare_methods(validation, metric = "rmse")
comparison$ranking[, c(
  "method", "rmse", "mae", "me", "finite_coverage",
  "support_coverage", "rank"
)]
```

```{r validation-plot, fig.alt="Observed versus predicted synthetic heads for IDW and TPS three-fold validation, with a dashed one-to-one reference line and open symbols for limited-support predictions."}
ps_validation_plot(validation, type = "observed_predicted")
legend("topleft", legend = c("IDW", "TPS", "1:1"),
       col = c(2, 3, 1), pch = c(16, 16, NA), lty = c(NA, NA, 2), bty = "n")
```

Candidate settings are evaluated within the resampling design. The small grid
below illustrates mechanics; it is not a comprehensive IDW calibration.

```{r tuning}
tuning <- ps_tune_interpolation(
  points,
  method = "IDW",
  candidates = data.frame(idw_power = c(1.5, 2)),
  inner_design = "kfold",
  inner_folds = 3,
  refit = FALSE,
  seed = 21
)
tuning$candidates[, c("candidate_id", "metric", "coverage", "rank", "selected")]
```

## 7. Inspect spatial dependence, direction, and external drift

An empirical variogram is a diagnostic summary whose stability depends on
network geometry, lag definitions, trend treatment, and pair counts. A fitted
variogram is not justified simply because software returns parameters.

```{r variogram}
empirical <- ps_variogram(points, directions = c(0, 90))
head(empirical$empirical[, c("np", "dist", "gamma", "dir.hor")])

anis_points <- ps_make_points(
  synthetic_anisotropic_points,
  "x", "y", "head", "point_id", "EPSG:26916"
)
anisotropy <- ps_anisotropy(anis_points, minimum_pairs = 5)
anisotropy$summary
```

```{r variogram-plot, fig.alt="Directional empirical semivariance points for north-south and east-west directions; point size increases with the number of contributing observation pairs."}
vtab <- empirical$empirical
direction_cols <- c("0" = "#176b87", "90" = "#c65d2e")
plot(vtab$dist, vtab$gamma,
     col = direction_cols[as.character(vtab$dir.hor)],
     pch = 19, cex = 0.7 + sqrt(vtab$np) / 4,
     xlab = "Lag distance (m)", ylab = "Semivariance",
     main = "Directional empirical variograms")
legend("topleft", names(direction_cols), col = direction_cols, pch = 19,
       title = "Direction (degrees)", bty = "n")
```

If `anisotropy$summary` reports insufficient usable directions, the correct
response is to record that limitation—not to force a ratio. External drift
similarly needs a defensible covariate that is available throughout the
prediction domain. This deterministic illustration creates a compatible land
surface from the same template, then fits UK with that declared trend.

```{r external-drift}
land <- fit$template
xy <- xyFromCell(land, seq_len(ncell(land)))
values(land) <- 185 - 0.0005 * (xy[, 1] - mean(xy[, 1]))
names(land) <- "land_surface"

drift_points <- points
land_at_points <- extract(land, drift_points)[[2]]
values(drift_points)$Z <- 0.7 * land_at_points + 39

external_drift <- ps_interpolate(
  drift_points,
  methods = "UK",
  template = fit$template,
  trend = Z ~ land_surface,
  covariates = list(land_surface = land),
  return = "result"
)
external_drift$method_parameters
```

Association with a drift covariate is not evidence of causality, and using a
post-outcome or unavailable future covariate would invalidate prediction.

## 8. Separate support, method disagreement, and uncertainty

These products answer different questions. Support summarizes observation
geometry and declared thresholds. An ensemble is a specified combination of
methods. Disagreement is descriptive method spread. Kriging standard error is
conditional on the fitted ordinary-kriging model.

```{r uncertainty-objects}
ensemble <- ps_surface_ensemble(fit$surfaces[c("IDW", "TPS")])
disagreement <- ps_method_disagreement(fit$surfaces[c("IDW", "TPS")])

ok_uncertainty <- ps_surface_uncertainty(
  ps_interpolate(points, methods = "OK", template = fit$template,
                 return = "result"),
  approach = "kriging_variance"
)

fit$support$summary
disagreement$summary
ok_uncertainty$summary
```

```{r support-uncertainty-map, fig.width=10, fig.height=4, fig.alt="Three synthetic maps showing prediction-support class, absolute IDW versus TPS head disagreement, and ordinary-kriging prediction standard error; each describes a different limitation."}
par(mfrow = c(1, 3), mar = c(2.5, 2.5, 3, 1))
plot(fit$support$rasters[["support_class_code"]],
     col = c("#d8efe8", "#f3cf72", "#d77c63"),
     main = "Support class", axes = FALSE)
plot(disagreement$rasters[["range"]], col = hcl.colors(64, "YlOrRd"),
     main = "IDW–TPS range", axes = FALSE)
plot(ok_uncertainty$standard_deviation, col = hcl.colors(64, "YlOrRd"),
     main = "OK standard error", axes = FALSE)
```

None of these maps is a universal confidence surface. In particular, method
spread can be small when methods share the same bias, and geometric support
classes do not establish that a mapped value is correct.

## 9. Classify contours and check model-gradient symbols

The support thresholds below are analyst-declared distances. The classes mean
"meets these criteria," not "statistically confident." Arrows are derived from
the IDW raster and shortened when necessary so their endpoints remain within
finite raster support.

```{r contour-arrow-products}
contours <- ps_contours(fit$surfaces$IDW, interval = 1)
supported_contours <- ps_contour_support(
  contours,
  support = fit$support,
  supported_distance = 500,
  approximate_distance = 1000
)
flow <- ps_flow_arrows(
  fit$surfaces$IDW,
  res_factor = 3,
  scale = 40,
  endpoint_action = "shorten"
)
supported_contours$summary
flow$validation_summary
```

```{r contour-arrow-map, fig.width=10, fig.height=4.5, fig.alt="Two panels: contours drawn solid, dashed, or dotted by user-defined prediction-support criteria; and an IDW head surface with contours and checked arrows pointing toward lower modeled head."}
par(mfrow = c(1, 2), mar = c(2.5, 2.5, 3, 1))
plot(supported_contours, legend_position = NULL,
     main = "User-defined contour support")
plot(points, add = TRUE, pch = 21, bg = "white", cex = 0.55)
legend("bottomright", c("supported", "approximate", "unsupported"),
       lty = c(1, 2, 3), title = "Criteria", cex = 0.78, bty = "n")

plot(fit$surfaces$IDW, col = head_cols,
     main = "Modeled gradient direction", axes = FALSE)
plot(contours, add = TRUE, col = "#52646d", lwd = 0.8)
draw_flow(flow$arrows)
plot(points, add = TRUE, pch = 21, bg = "white", cex = 0.5)
```

Arrow direction indicates decreasing modeled head. Arrow length is a display
scale, not pore-water velocity, Darcy flux, travel time, or a particle path.

## 10. Compare events and state vertical-gradient sign

The event comparison pairs wells by identifier and retains measured and
modeled differences. The signed raster is autumn minus spring because autumn
is passed as event B.

```{r temporal-products}
spring_ps <- ps_make_points(spring, "x", "y", "head", "well_id", "EPSG:26916")
autumn_ps <- ps_make_points(autumn, "x", "y", "head", "well_id", "EPSG:26916")

change <- ps_head_change(
  spring_ps, autumn_ps,
  pair_by = "well_id",
  method = "IDW",
  grid_res = 350
)
change$summary

vertical <- ps_vertical_gradient(
  upper_head = 10,
  lower_head = 12,
  upper_elevation = 100,
  lower_elevation = 80,
  positive = "upward"
)
vertical$summary
```

```{r change-map, fig.alt="Synthetic modeled autumn-minus-spring head change raster colored from blue for negative change to red for positive change, with paired event wells overlaid."}
change_limit <- max(abs(minmax(change$modeled_change)), na.rm = TRUE)
plot(change$modeled_change, col = diff_cols,
     range = c(-change_limit, change_limit),
     main = "Modeled head change: autumn minus spring")
plot(spring_ps, add = TRUE, pch = 21, bg = "white", cex = 0.55)
```

Modeled head change is not storage change. Converting change in head to change
in storage requires appropriate storage properties, boundaries, aquifer
geometry, and a defensible water-budget interpretation. Vertical-gradient
sign likewise has meaning only when the numerator, denominator, reference
elevations, interval pairing, and positive convention are explicit.

## 11. Review influence, thinning, candidates, and settings

Network diagnostics describe consequences under their declared method and
scenario. They do not discover a uniquely optimal monitoring network.

```{r network-analysis}
influence <- ps_well_influence(
  points[1:7], method = "IDW", grid_res = 400
)
thinning <- ps_network_thinning(
  points[1:12], retain = 0.75, repeats = 2,
  method = "IDW", grid_res = 400, seed = 6
)

candidate_points <- vect(
  subset(synthetic_candidate_sites, !excluded),
  geom = c("x", "y"), crs = "EPSG:26916"
)
candidate_design <- ps_candidate_network(
  points,
  candidate_points,
  objective = "spatial_coverage",
  n_select = 2
)

sensitivity <- ps_surface_sensitivity(
  points[1:14],
  method = "IDW",
  scenarios = data.frame(idw_power = c(1.5, 2), grid_res = 350),
  reference = 1,
  seed = 21
)

head(influence$influence[order(-influence$influence$rmse_difference),
                         c("well_id", "rmse_difference", "held_out_residual")])
thinning$summary
candidate_design$candidate_scores
sensitivity$comparisons
```

```{r network-plot, fig.width=10, fig.height=4.5, fig.alt="Two network-review panels: leave-one-well surface RMSE difference by synthetic well and candidate sites around the existing monitoring network, with selected candidates highlighted."}
par(mfrow = c(1, 2), mar = c(7, 4, 3, 1))
ord <- order(influence$influence$rmse_difference, decreasing = TRUE)
barplot(influence$influence$rmse_difference[ord],
        names.arg = influence$influence$well_id[ord], las = 2,
        col = "#4f9daf", border = NA,
        ylab = "Leave-one-well surface RMSE difference",
        main = "Influence under IDW")

plot(points, pch = 21, bg = "#176b87", cex = 0.75,
     main = "Existing and candidate locations")
plot(candidate_points, add = TRUE, pch = 24, bg = "#f3cf72", cex = 1.1)
chosen_indices <- as.integer(sub(
  "candidate_0+", "", candidate_design$selected_sequence$candidate_id
))
chosen <- candidate_points[chosen_indices]
plot(chosen, add = TRUE, pch = 24, bg = "#d65f4b", cex = 1.35)
legend("bottomleft", c("existing", "candidate", "selected"),
       pch = c(21, 24, 24), pt.bg = c("#176b87", "#f3cf72", "#d65f4b"),
       bty = "n")
```

Candidate ranking must also respect access, ownership, utilities, construction
feasibility, screened interval, sampling purpose, budget, and permitting. A
high algorithmic score is not authorization to drill.

## 12. Split regions and make depth, profile, and section products

Separate hydrogeologic regions can be interpolated only when their boundaries
and group assignments are defensible. The synthetic regions illustrate the
API; they do not imply that a statistical discontinuity has been established.

```{r regions-depth-profile}
regions <- vect(synthetic_regions, geom = "wkt", crs = "EPSG:26916")
regional <- ps_interpolate_regions(
  points, regions, region_id = "region_id",
  methods = "IDW", grid_res = 350
)

depth <- ps_depth_to_water_surface(
  fit$surfaces$IDW, land, surface_type = "potentiometric"
)

transect <- vect(synthetic_transect, geom = "wkt", crs = "EPSG:26916")
profile <- ps_surface_profile(
  transect,
  surfaces = list(head = fit$surfaces$IDW, land = land),
  n = 12
)
section <- ps_cross_section(
  transect,
  head_surface = fit$surfaces$IDW,
  land_surface = land,
  step = 300,
  vertical_exaggeration = 3
)

regional$summary
depth$summary
profile$summary
section$summary
```

```{r profile-section-plot, fig.width=10, fig.height=4.5, fig.alt="Two transect panels: head and land-surface elevation plotted against chainage, and a cross-section view of the same profiles with three-times vertical exaggeration declared in the title."}
par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))
ptab <- profile$profile
plot(ptab$chainage, ptab$land, type = "l", lty = 2, lwd = 2,
     col = "#6b5946", xlab = "Chainage (m)", ylab = "Elevation (m)",
     main = "Surface profile")
lines(ptab$chainage, ptab$head, lwd = 2, col = "#176b87")
legend("bottomleft", c("land", "modeled head"),
       lty = c(2, 1), lwd = 2, col = c("#6b5946", "#176b87"), bty = "n")

plot(section, lwd = 2, col = "#176b87",
     main = "Plot-ready section (3× vertical exaggeration)")
legend("bottomleft", c("modeled head", "land"), lty = c(1, 2),
       col = c("#176b87", "black"), bty = "n")
```

Depth is meaningful only when land and head surfaces share compatible
horizontal and vertical references and units. The cross-section samples
surfaces along a line; it does not invent hydrostratigraphy or become a
three-dimensional numerical model.

## 13. Export auditable products and record the interpretation

Exports should preserve method names, units, CRS, settings, support criteria,
conditions, and software versions. This chunk writes a raster, contours,
quicklook, contour-support files, and an open QGIS style to a temporary folder,
then removes the folder after listing its products.

```{r export}
output_dir <- file.path(tempdir(), "potentiomap-complete-example")
if (dir.exists(output_dir)) unlink(output_dir, recursive = TRUE)
dir.create(output_dir)

surface_files <- ps_export_surfaces(
  list(IDW = fit$surfaces$IDW),
  out_dir = output_dir,
  out_stub = "synthetic_head",
  contour_interval = 1,
  points = points,
  support = fit$support,
  diagnostics = fit$diagnostics,
  write_support = TRUE,
  write_diagnostics = TRUE,
  vector_format = "gpkg"
)
support_files <- ps_export_contour_support(
  supported_contours,
  out_dir = output_dir,
  out_stub = "synthetic_contour_support",
  vector_format = "gpkg"
)
style_file <- ps_export_style(
  fit$surfaces$IDW,
  file = file.path(output_dir, "synthetic_head.qml"),
  format = "qml",
  layer_type = "head_raster",
  units = "m"
)

data.frame(file = sort(list.files(output_dir)))
unlink(output_dir, recursive = TRUE)
```

A release-quality analysis should close with a plain-language record of what
the map represents, which observations and screens were used, when they were
measured, which prediction task was validated, where support is limited, which
uncertainty assumptions apply, and which decisions the products cannot support.

<div class="pm-method-note">
Reproducibility checklist: package version 0.2.0; fixed seeds shown in code; installed synthetic inputs only; EPSG:26916 projected coordinates; metre-based synthetic heads and elevations; explicit event, screen, validation, support, and sign conventions; temporary exports; no external services or repository development scripts.
</div>

```{r session-info}
sessionInfo()
```
