---
title: "Contours and surface smoothing"
description: "Regular and explicit contours plus mean, median, one-pass, and multi-pass focal smoothing with scientific cautions."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Contours and surface smoothing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 9, fig.height = 5, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
cols <- hcl.colors(64, "RdYlBu", rev = TRUE)
data("synthetic_wells", package = "potentiomap")
points <- ps_make_points(synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916")
surface <- ps_interpolate(points, "Z", "TPS", grid_res = 75,
                          mask = ps_sample_aoi(), padding = 0)$TPS
```

## Regular intervals and explicit levels

```{r contour-types}
regular <- ps_contours(surface, interval = 1)
explicit <- ps_contours(surface, levels = c(166, 168, 170))
data.frame(
  contour_set = c("1-unit interval", "explicit 166, 168, 170"),
  line_features = c(nrow(regular), nrow(explicit)),
  attributes = c(paste(names(regular), collapse = ", "),
                 paste(names(explicit), collapse = ", "))
)
```

```{r raw-contours, fig.alt="Synthetic TPS surface shaded from cool low heads to warm high heads, beside a contour-only map with one-unit lines and monitoring locations."}
par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
plot(surface, col = cols, main = "Raw modeled surface")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
plot(ps_sample_aoi(), col = "#f7fafb", border = "#52646d", main = "Raw 1-unit contours")
plot(regular, add = TRUE, col = "#176b87", lwd = 1.2)
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
```

## Mean and median smoothing

```{r smooth-variants}
mean_one <- ps_smooth_surface(surface, window_size = 5, method = "mean", iterations = 1)
median_one <- ps_smooth_surface(surface, window_size = 5, method = "median", iterations = 1)
mean_three <- ps_smooth_surface(surface, window_size = 5, method = "mean", iterations = 3)

data.frame(
  surface = c("original", "mean one pass", "median one pass", "mean three passes"),
  minimum = vapply(list(surface, mean_one, median_one, mean_three), function(x) minmax(x)[1], numeric(1)),
  maximum = vapply(list(surface, mean_one, median_one, mean_three), function(x) minmax(x)[2], numeric(1)),
  standard_deviation = vapply(list(surface, mean_one, median_one, mean_three),
                              function(x) global(x, "sd", na.rm = TRUE)[[1]], numeric(1))
)
```

```{r smooth-surfaces, fig.width=10, fig.height=8, fig.alt="Four synthetic modeled surfaces on a common head scale: original TPS, one-pass mean smoothing, one-pass median smoothing, and three-pass mean smoothing."}
all_surfaces <- list(original = surface, mean_one = mean_one,
                     median_one = median_one, mean_three = mean_three)
shared <- range(unlist(lapply(all_surfaces, minmax)), finite = TRUE)
par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))
for (nm in names(all_surfaces)) {
  plot(all_surfaces[[nm]], col = cols, range = shared,
       main = gsub("_", " ", nm))
  plot(points, add = TRUE, pch = 21, bg = "white", cex = .5)
}
```

```{r smoothed-contours, fig.alt="Original and mean-smoothed TPS surfaces displayed side by side with one-unit contours, showing small changes in contour shape after smoothing."}
mean_contours <- ps_contours(mean_one, interval = 1)
par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
plot(surface, col = cols, range = shared, main = "Original surface and contours")
plot(regular, add = TRUE, col = "#17252d")
plot(mean_one, col = cols, range = shared, main = "Mean-smoothed surface and contours")
plot(mean_contours, add = TRUE, col = "#17252d")
```

```{r preserve-footprint}
data.frame(
  object = c("original", "smoothed"),
  missing_cells = c(global(is.na(surface), "sum")[[1]],
                    global(is.na(mean_one), "sum")[[1]]),
  same_geometry = c(TRUE, compareGeom(surface, mean_one, stopOnError = FALSE))
)
```

<div class="pm-science-note">
Smoothing changes the modeled surface; it is not an accuracy correction. Retain the unsmoothed result, record the method, window, and number of passes, and review whether local variation is being obscured. Contours derived from a smoothed raster describe that smoothed model, not the original interpolation.
</div>
