---
title: "Custom interpolation functions"
description: "Implement and validate the released custom-method signature without using package internals."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom interpolation functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 8, fig.height = 6, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
```

A custom function is called exactly as `fun(points, template, grid)`. It must
return either a `SpatRaster` matching the template or one numeric value per
template cell. The following plane is intentionally simple: it demonstrates
the interface, not a scientific recommendation.

```{r define-custom}
planar_method <- function(points, template, grid) {
  stopifnot(
    inherits(points, "SpatVector"),
    inherits(template, "SpatRaster"),
    is.data.frame(grid),
    identical(names(grid), c("X", "Y")),
    nrow(grid) == ncell(template),
    "Z" %in% names(values(points))
  )

  xy <- as.data.frame(crds(points))
  names(xy) <- c("X", "Y")
  xy$Z <- values(points)$Z
  if (nrow(xy) < 3 || any(!is.finite(xy$Z))) {
    stop("The planar example needs at least three finite observations.")
  }

  fit <- lm(Z ~ X + Y, data = xy)
  prediction <- predict(fit, newdata = grid)
  stopifnot(is.numeric(prediction), length(prediction) == ncell(template))
  prediction
}
```

```{r run-custom}
data("synthetic_wells", package = "potentiomap")
points <- ps_make_points(
  synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916"
)

surface <- ps_interpolate(
  points,
  methods = "planar",
  custom_methods = list(planar = planar_method),
  grid_res = 100,
  mask = ps_sample_aoi(),
  padding = 0
)$planar

data.frame(
  class = class(surface)[1],
  rows = nrow(surface), columns = ncol(surface),
  x_resolution_m = res(surface)[1],
  finite_cells = global(surface, "notNA")[[1]]
)
```

```{r custom-map, fig.alt="A simple planar custom interpolation of synthetic groundwater heads, shaded blue at lower modeled heads and red at higher modeled heads, with monitoring wells overlaid."}
plot(surface, col = hcl.colors(64, "RdYlBu", rev = TRUE),
     main = "Custom planar example — not a recommended model")
plot(points, add = TRUE, pch = 21, bg = "white")
```

```{r geometry-check}
expected <- rast(crs = crs(points), ext = ext(ps_sample_aoi()), resolution = 100)
data.frame(
  same_crs = same.crs(surface, expected),
  x_resolution_m = res(surface)[1],
  y_resolution_m = res(surface)[2],
  masked = global(is.na(surface), "sum")[[1]] > 0
)
```

For a production custom method, add method-specific diagnostics, parameter
records, unit tests, and validation. Do not rely on `potentiomap:::` internals;
the documented function signature is the stable public boundary.
