---
title: "Exporting GIS-ready products"
description: "Write, inventory, read back, and validate GeoTIFF, contour, quicklook, arrow, tip, and base products."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exporting GIS-ready products}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 8, fig.height = 6, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
data("synthetic_wells", package = "potentiomap")
points <- ps_make_points(synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916")
surfaces <- ps_interpolate(points, methods = c("TPS", "IDW"), grid_res = 100,
                           mask = ps_sample_aoi(), padding = 0)
```

The package exports standard spatial files. The example writes to an R session
temporary directory, then prints only portable file names so local paths never
become part of a project record.

## Export surfaces, contours, and quicklooks

```{r export-surfaces}
output_dir <- file.path(tempdir(), "potentiomap-gis-example")
inventory <- ps_export_surfaces(
  surfaces,
  out_dir = output_dir,
  out_stub = "synthetic",
  contour_interval = 1,
  points = points
)

portable_inventory <- transform(
  inventory,
  raster = basename(raster),
  contours = basename(contours),
  quicklook = basename(quicklook)
)
portable_inventory
```

GeoPackage is the recommended vector format because it preserves field names
and related attributes in one file. Shapefile remains available when required;
preserve every sidecar file when transferring it.

## Read the products back

```{r read-back}
surface_back <- rast(inventory$raster[1])
contours_back <- vect(inventory$contours[1])

data.frame(
  check = c("raster CRS", "contour CRS", "raster dimensions", "contour attributes", "contour features"),
  result = c(
    crs(surface_back, proj = TRUE),
    crs(contours_back, proj = TRUE),
    paste(dim(surface_back), collapse = " × "),
    paste(names(contours_back), collapse = ", "),
    nrow(contours_back)
  )
)
```

```{r read-back-map, fig.alt="GeoTIFF TPS surface read back from disk, shaded blue at lower synthetic heads and red at higher heads, with the exported contour shapefile overlaid."}
plot(surface_back, col = hcl.colors(64, "RdYlBu", rev = TRUE),
     main = "Read-back TPS GeoTIFF and contour vector")
plot(contours_back, add = TRUE, col = "#17252d", lwd = 1)
plot(points, add = TRUE, pch = 21, bg = "white", cex = .65)
```

## Export hydraulic-gradient products

```{r export-flow}
flow <- ps_flow_arrows(
  surfaces$TPS,
  res_factor = 8,
  scale = 75,
  min_gradient = 1e-5,
  out_dir = output_dir,
  out_stub = "synthetic_TPS",
  endpoint_action = "shorten"
)
tips_file <- file.path(output_dir, "synthetic_TPS_arrow_tips.gpkg")
bases_file <- file.path(output_dir, "synthetic_TPS_arrow_bases.gpkg")
tips <- ps_arrow_vertices(flow$arrows, "last", tips_file)
bases <- ps_arrow_vertices(flow$arrows, "first", bases_file)

data.frame(
  product = c("gradient GeoTIFF", "sample points", "arrow lines", "arrow tips", "arrow bases"),
  file = c("synthetic_TPS_hgrad.tif", "synthetic_TPS_hgrad_points.shp",
           "synthetic_TPS_hgrad_arrows.shp", basename(tips_file), basename(bases_file)),
  features_or_layers = c(nlyr(flow$raster), nrow(flow$points), nrow(flow$arrows), nrow(tips), nrow(bases))
)
```

## Validate geometry and attributes

```{r export-validation}
gradient_back <- rast(file.path(output_dir, "synthetic_TPS_hgrad.tif"))
arrow_back <- vect(file.path(output_dir, "synthetic_TPS_hgrad_arrows.shp"))
data.frame(
  check = c("gradient layers", "gradient CRS matches surface", "arrow CRS matches surface",
            "arrow attributes", "tip count equals arrow count"),
  pass = c(
    identical(names(gradient_back), c("gwe", "igrad", "aspect")),
    same.crs(gradient_back, surfaces$TPS),
    same.crs(arrow_back, surfaces$TPS),
    all(c("gwe", "igrad", "aspect") %in% names(arrow_back)),
    nrow(tips) == nrow(arrow_back)
  )
)
```

The tables and read-back map above are generated during the website build.
Use the displayed code with a project-specific output directory to create the
same inventory under the installed package version.

<div class="pm-science-note">
Exported files preserve the modeled products, not their validity. Keep the monitoring data, aquifer selection, vertical datum, interpolation parameters, mask, software version, and review record with every deliverable.
</div>
