---
title: "Input formats and coordinate systems"
description: "Prepare data-frame, sf, and terra point inputs; distinguish CRS assignment from transformation; and diagnose common spatial mistakes."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input formats and coordinate systems}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 8, fig.height = 5, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
data("synthetic_wells", package = "potentiomap")
```

Distance-based interpolation requires meaningful distances. A projected CRS is
generally appropriate for local potentiometric mapping because grid resolution,
padding, search distances, and gradients are then expressed in known linear map
units.

## Three supported point inputs

```{r three-inputs}
from_table <- ps_make_points(
  synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916"
)
from_sf <- ps_make_points(
  sf::st_as_sf(synthetic_wells, coords = c("x", "y"), crs = 26916),
  value = "gw_elevation", name_col = "well_id"
)
from_terra <- ps_make_points(
  vect(synthetic_wells, geom = c("x", "y"), crs = "EPSG:26916"),
  value = "gw_elevation", name_col = "well_id"
)
data.frame(
  input = c("data frame", "sf points", "terra SpatVector"),
  rows = c(nrow(from_table), nrow(from_sf), nrow(from_terra)),
  projected = c(!is.lonlat(from_table), !is.lonlat(from_sf), !is.lonlat(from_terra)),
  crs = rep(crs(from_table, proj = TRUE), 3)
)
```

## Assigning is not transforming

Assign a CRS only when the coordinate numbers already use that reference
system. Transform when converting known coordinates to another CRS.

```{r assignment-transformation}
lonlat <- data.frame(
  id = paste0("W", 1:6),
  longitude = c(-92.99, -92.98, -92.97, -92.96, -92.95, -92.94),
  latitude = c(34.50, 34.52, 34.51, 34.54, 34.53, 34.55),
  head_ft = c(510, 505, 500, 496, 492, 487)
)
geographic <- ps_make_points(
  lonlat, "longitude", "latitude", "head_ft", "id", "EPSG:4326"
)
projected <- project(geographic, "EPSG:26915")

data.frame(
  object = c("assigned geographic", "transformed projected"),
  is_longitude_latitude = c(is.lonlat(geographic), is.lonlat(projected)),
  x_range = c(paste(round(range(crds(geographic)[, 1]), 3), collapse = " to "),
              paste(round(range(crds(projected)[, 1]), 1), collapse = " to "))
)
```

Groundwater points may arrive in geographic coordinates, but map-distance
support classification rejects longitude/latitude degrees. Transform to a
suitable projected CRS before distance-based interpolation or classification:

```{r geographic-check}
if (is.lonlat(geographic)) {
  message("Geographic coordinates detected; transform before distance-based interpolation.")
}
```

This explicit check also makes horizontal units visible in the analysis
record.

## Actual missing-CRS error

```{r missing-crs-error}
missing_crs_message <- tryCatch(
  ps_make_points(synthetic_wells, "x", "y", "gw_elevation", "well_id"),
  error = function(e) conditionMessage(e)
)
missing_crs_message
```

## Inspect extent, units, and AOI overlap

```{r spatial-checks}
aoi <- ps_sample_aoi()
point_extent <- ext(from_table)
aoi_extent <- ext(aoi)
inside <- relate(from_table, aoi, "within")
data.frame(
  check = c("projected coordinates", "same CRS", "points inside AOI", "linear unit"),
  result = c(!is.lonlat(from_table), same.crs(from_table, aoi),
             paste(sum(inside), "of", length(inside)), "metre")
)
```

```{r extent-map, fig.alt="Synthetic monitoring wells shown inside a pale area-of-interest polygon in projected UTM coordinates, with axes in metres."}
plot(aoi, col = "#e8f3f6", border = "#176b87", lwd = 2,
     main = "Check points against the AOI")
plot(from_table, add = TRUE, pch = 21, bg = "#f4b183")
```

Before interpolation, verify the horizontal datum and units, vertical datum and
head units, coordinate ranges, transformation pipeline, AOI overlap, and whether
the projected CRS is suitable for the project location. Mismatched raster and
vector CRSs should be transformed deliberately, not relabeled.
