---
title: "Comparing TPS, IDW, ordinary kriging, and universal kriging"
description: "Four released interpolation methods on one projected grid, with common scales, contours, and difference rasters."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing TPS, IDW, ordinary kriging, and universal kriging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 9, fig.height = 7, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
head_cols <- hcl.colors(64, "RdYlBu", rev = TRUE)
diff_cols <- hcl.colors(64, "Blue-Red 3", rev = TRUE)
```

Interpolation choice should reflect monitoring-network geometry, site
understanding, assumptions, and validation—not visual smoothness alone. This
comparison holds the observations, projected CRS, area of interest, 100-metre
grid, raster geometry, and mask constant.

## One common template

```{r common-inputs}
data("synthetic_wells", package = "potentiomap")
points <- ps_make_points(
  synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916"
)
aoi <- ps_sample_aoi()

template <- rast(crs = crs(points), ext = ext(aoi), resolution = 100)
template
```

## Fit all four released methods

```{r fit-methods}
surfaces <- ps_interpolate(
  points,
  methods = c("TPS", "IDW", "OK", "UK"),
  template = template,
  mask = aoi,
  idw_power = 2,
  idw_nmax = 15,
  tps_lambda = NULL,
  kr_auto_cutoff = TRUE
)
names(surfaces)
```

The displayed TPS or variogram warnings are part of the actual fit. A warning
does not establish that a model is fit for a consequential decision; inspect
diagnostics and validate against withheld observations when that decision
requires it.

```{r method-summaries}
do.call(rbind, lapply(names(surfaces), function(method) {
  r <- surfaces[[method]]
  data.frame(
    method = method,
    rows = nrow(r), columns = ncol(r),
    x_resolution_m = res(r)[1], y_resolution_m = res(r)[2],
    minimum = minmax(r)[1], maximum = minmax(r)[2]
  )
}))
```

## Surfaces and contours on one head scale

```{r method-maps, fig.width=10, fig.height=9, fig.alt="Four synthetic potentiometric-surface maps for TPS, IDW, ordinary kriging, and universal kriging using one blue-low to red-high head scale, identical grid geometry, one-unit contours, and the same monitoring points."}
common_range <- range(unlist(lapply(surfaces, minmax)), finite = TRUE)
contours <- lapply(surfaces, ps_contours, interval = 1)

par(mfrow = c(2, 2), mar = c(3, 3, 3, 1))
for (method in names(surfaces)) {
  plot(surfaces[[method]], col = head_cols, range = common_range, main = method)
  plot(contours[[method]], add = TRUE, col = "#17252d", lwd = .9)
  plot(points, add = TRUE, pch = 21, bg = "white", cex = .55)
}
```

TPS uses a thin-plate spline and its smoothing parameter; IDW is controlled by
power and neighbor selection. OK estimates a constant-mean geostatistical
surface, while UK uses the released quadratic drift specification. The package
fits an empirical variogram for the kriging methods.

## Differences relative to TPS

```{r difference-summary}
differences <- lapply(surfaces[c("IDW", "OK", "UK")], function(x) x - surfaces$TPS)
do.call(rbind, lapply(names(differences), function(method) {
  x <- differences[[method]]
  data.frame(
    comparison = paste(method, "minus TPS"),
    mean_difference = global(x, "mean", na.rm = TRUE)[[1]],
    mean_absolute_difference = global(abs(x), "mean", na.rm = TRUE)[[1]],
    maximum_absolute_difference = global(abs(x), "max", na.rm = TRUE)[[1]]
  )
}))
```

```{r difference-maps, fig.width=10, fig.height=4, fig.alt="Three difference rasters showing IDW minus TPS, ordinary kriging minus TPS, and universal kriging minus TPS with a centered blue-white-red difference scale."}
limit <- max(abs(unlist(lapply(differences, minmax))), na.rm = TRUE)
par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))
for (method in names(differences)) {
  plot(differences[[method]], col = diff_cols, range = c(-limit, limit),
       main = paste(method, "− TPS"))
  plot(points, add = TRUE, pch = 21, bg = "white", cex = .45)
}
```

No method is universally best. Compare results against hydrogeologic
boundaries, screened intervals, surface-water controls, network spacing, and an
appropriate validation strategy. Increased smoothness or visual detail is not
evidence of increased accuracy.
