---
title: "Interpolation parameters and common grid geometry"
description: "See how resolution, padding, masks, IDW controls, TPS smoothing, kriging lags, and templates affect the modeled product."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolation parameters and common grid geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 9, fig.height = 5, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
head_cols <- hcl.colors(64, "RdYlBu", rev = TRUE)
data("synthetic_wells", package = "potentiomap")
points <- ps_make_points(synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916")
aoi <- ps_sample_aoi()
```

Parameters define the numerical and spatial problem. Record them with each
monitoring event. More cells or a smoother image do not create new observations.

## Coarse and fine resolution

```{r resolution-build}
coarse <- ps_interpolate(points, "Z", "TPS", grid_res = 200, mask = aoi, padding = 0)$TPS
fine <- ps_interpolate(points, "Z", "TPS", grid_res = 50, mask = aoi, padding = 0)$TPS
data.frame(
  grid = c("coarse", "fine"),
  cell_size_m = c(res(coarse)[1], res(fine)[1]),
  cells = c(ncell(coarse), ncell(fine)),
  cells_with_values = c(global(coarse, "notNA")[[1]], global(fine, "notNA")[[1]])
)
```

```{r resolution-plot, fig.alt="Side-by-side synthetic TPS surfaces at 200-metre and 50-metre cell sizes; both show the same broad head pattern while the finer raster contains more display detail."}
shared <- range(minmax(coarse), minmax(fine))
par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
plot(coarse, col = head_cols, range = shared, main = "200 m cells")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
plot(fine, col = head_cols, range = shared, main = "50 m cells")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
```

## Padding and the AOI mask

Padding expands the prediction extent around the point or AOI extent. A mask
then limits the published footprint to the analysis polygon.

```{r padding-build}
pad_small <- ps_interpolate(points, methods = "IDW", grid_res = 100, padding = 100)$IDW
pad_large <- ps_interpolate(points, methods = "IDW", grid_res = 100, padding = 700)$IDW
rbind(
  small_padding = as.vector(ext(pad_small)),
  large_padding = as.vector(ext(pad_large))
)
```

```{r padding-plot, fig.alt="Side-by-side unmasked IDW rasters with 100-metre and 700-metre padding, showing how padding changes prediction extent beyond the monitoring network."}
par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
plot(pad_small, col = head_cols, main = "100 m padding")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
plot(pad_large, col = head_cols, main = "700 m padding")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
```

Predictions outside the support of the monitoring network are extrapolations,
regardless of whether a rectangular raster is displayed.

## IDW power and neighbor limit

```{r idw-power}
idw_1 <- ps_interpolate(points, methods = "IDW", grid_res = 75, mask = aoi,
                        padding = 0, idw_power = 1, idw_nmax = 15)$IDW
idw_3 <- ps_interpolate(points, methods = "IDW", grid_res = 75, mask = aoi,
                        padding = 0, idw_power = 3, idw_nmax = 8)$IDW
data.frame(
  setting = c("power 1, nmax 15", "power 3, nmax 8"),
  minimum = c(minmax(idw_1)[1], minmax(idw_3)[1]),
  maximum = c(minmax(idw_1)[2], minmax(idw_3)[2])
)
```

```{r idw-power-plot, fig.alt="Two synthetic IDW surfaces on a common scale; power one with fifteen neighbors is more regionally blended, while power three with eight neighbors emphasizes nearby wells more strongly."}
shared <- range(minmax(idw_1), minmax(idw_3))
par(mfrow = c(1, 2), mar = c(3, 3, 3, 1))
plot(idw_1, col = head_cols, range = shared, main = "IDW power 1; nmax 15")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
plot(idw_3, col = head_cols, range = shared, main = "IDW power 3; nmax 8")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .6)
```

## TPS and kriging controls

```{r tps-kriging-controls}
tps_fixed <- ps_interpolate(
  points, methods = "TPS", grid_res = 100, mask = aoi, padding = 0,
  tps_lambda = 0.001
)$TPS

ok_manual <- ps_interpolate(
  points, methods = "OK", grid_res = 100, mask = aoi, padding = 0,
  kr_auto_cutoff = FALSE, kr_cutoff = 1400, kr_width = 100
)$OK

data.frame(
  surface = c("TPS lambda 0.001", "OK cutoff 1400 m; width 100 m"),
  minimum = c(minmax(tps_fixed)[1], minmax(ok_manual)[1]),
  maximum = c(minmax(tps_fixed)[2], minmax(ok_manual)[2])
)
```

`tps_lambda = NULL` delegates smoothing selection to `fields::Tps()` by GCV.
Kriging lag choices affect the empirical variogram and fit. Inspect warnings;
do not treat automatic or manually supplied settings as validation.

## Reuse one template

```{r common-template}
template <- rast(crs = crs(points), ext = ext(aoi), resolution = 100)
same_grid <- ps_interpolate(
  points, methods = c("TPS", "IDW"), template = template, mask = aoi
)
compareGeom(same_grid$TPS, same_grid$IDW, stopOnError = FALSE)
```

Common geometry is essential for cell-by-cell differences and repeated-event
comparisons. Save the template or its CRS, extent, resolution, origin, and mask
with the project record.
