---
title: "Five-minute quick start"
description: "A copy-paste workflow from released synthetic groundwater observations to a surface, contours, quicklook, and hydraulic-gradient arrows."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Five-minute quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 8, fig.height = 6, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
palette_head <- hcl.colors(64, "RdYlBu", rev = TRUE)
```

CRAN currently serves `potentiomap` 0.1.0, while this website documents 0.2.0.
Install the website version from GitHub when you need the full workflow below:

```{r install-website-version, eval=FALSE}
install.packages("pak")
pak::pak("el-cordero/potentiomap")
```

The example uses released synthetic data in EPSG:26916. The head values are
synthetic elevation units rather than field measurements.

## Prepare the observations

```{r prepare-points}
data("synthetic_wells", package = "potentiomap")

points <- ps_make_points(
  synthetic_wells,
  x = "x", y = "y",
  value = "gw_elevation",
  name_col = "well_id",
  crs = "EPSG:26916"
)

head(values(points)[, c("Name", "Z")])
```

`Z` is the standardized head field and `Name` identifies each observation.

## Build the surface and contours

```{r interpolate}
aoi <- ps_sample_aoi()
surfaces <- ps_interpolate(
  points,
  methods = "TPS",
  grid_res = 100,
  mask = aoi,
  padding = 0
)
contours <- ps_contours(surfaces$TPS, interval = 1)

data.frame(
  output = names(surfaces),
  rows = nrow(surfaces$TPS),
  columns = ncol(surfaces$TPS),
  cells_with_values = global(surfaces$TPS, "notNA")[[1]],
  contour_features = nrow(contours)
)
```

## Preserve the native quicklook

```{r quicklook, fig.alt="Native potentiomap quicklook of a synthetic TPS surface, with blue and white one-unit contours and labeled monitoring wells."}
ps_quicklook(
  surfaces$TPS,
  contours = contours,
  points = points,
  title = "Synthetic TPS potentiometric surface"
)
```

## Derive inferred hydraulic-gradient arrows

```{r flow-objects}
flow <- ps_flow_arrows(
  surfaces$TPS,
  res_factor = 8,
  scale = 75,
  min_gradient = 1e-5,
  endpoint_action = "shorten"
)
tips <- ps_arrow_vertices(flow$arrows, which = "last")
bases <- ps_arrow_vertices(flow$arrows, which = "first")

data.frame(
  gradient_layers = paste(names(flow$raster), collapse = ", "),
  arrow_sample_points = nrow(flow$points),
  arrow_lines = nrow(flow$arrows),
  tips = nrow(tips),
  bases = nrow(bases)
)
```

```{r flow-map, fig.alt="Synthetic TPS potentiometric surface with dark arrows pointing generally from warmer high modeled heads toward cooler low modeled heads."}
plot(surfaces$TPS, col = palette_head, main = "Inferred hydraulic-gradient direction")
plot(contours, add = TRUE, col = "#52646d", lwd = 1)
bxy <- crds(bases)
txy <- crds(tips)
arrows(bxy[, 1], bxy[, 2], txy[, 1], txy[, 2], length = 0.055, col = "#111111", lwd = 1.2)
plot(points, add = TRUE, pch = 21, bg = "white", cex = 0.75)
```

<div class="pm-science-note">
These lines represent direction derived from the interpolated head surface. Their lengths are cartographically scaled. They are not groundwater velocities, travel times, particle paths, or contaminant-transport trajectories.
</div>

Continue with [preparing observations](preparing-observations.html),
[comparing interpolation methods](interpolation-methods.html), or
[exporting GIS-ready products](exporting-products.html).
