---
title: "Public USGS groundwater-monitoring example"
description: "A reproducible TPS, contour, and inferred-gradient workflow for 36 Stanley Shale wells near Hot Springs, Arkansas, using official USGS data."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Public USGS groundwater-monitoring example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 9, fig.height = 6, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
cols <- hcl.colors(64, "RdYlBu", rev = TRUE)
data_candidates <- c(
  file.path("pkgdown", "assets", "downloads", "usgs_hot_springs_stanley_shale_2017.csv"),
  file.path("..", "pkgdown", "assets", "downloads", "usgs_hot_springs_stanley_shale_2017.csv"),
  file.path("..", "..", "pkgdown", "assets", "downloads", "usgs_hot_springs_stanley_shale_2017.csv")
)
data_path <- data_candidates[file.exists(data_candidates)][1]
stopifnot(length(data_path) == 1L, !is.na(data_path))
```

This example uses the official USGS data release
[10.5066/F7TD9WK0](https://doi.org/10.5066/F7TD9WK0), associated with
[Scientific Investigations Map 3444](https://doi.org/10.3133/sim3444). The
release describes 59 groundwater-level measurements collected in July and
August 2017 near the planned Highway 270 bypass east of Hot Springs, Arkansas.

The example demonstrates a reproducible software workflow. It does **not**
claim to reproduce the officially interpreted surface, and the official
contours are not treated as error-free truth.

## Retrieval and preparation record

The official ScienceBase aggregate download retrieved on 2026-07-13 contained a
zero-byte `.shp` geometry member. The official DBF attribute table was intact,
and authoritative NAD83 coordinates for all site identifiers were therefore
retrieved from the USGS NWIS site service. The prepared website CSV joins those
two official sources.

- ScienceBase aggregate ZIP SHA-256:
  `692543cc842082400be32ddc5b206fca1ac65d56b9e6358050c47cc0a538d528`
- Official DBF SHA-256:
  `73f0ec05d713fd886d7e801aae1ff9b3fea40c530a4eebf41ef96a611120ae59`
- Source release publication date: 2019-10-11
- Groundwater-level altitude: feet above NAVD 88
- Source longitude/latitude datum: NAD83

The map is restricted to the 36 records whose documented local aquifer code is
`330STNL` (Stanley Shale). This avoids mixing the five local aquifer codes in a
single software demonstration.

```{r inspect-public-data}
usgs <- read.csv(data_path, colClasses = c(site_id = "character"))
data.frame(
  records = nrow(usgs),
  local_aquifer_code = paste(unique(usgs$local_aquifer_code), collapse = ", "),
  measurement_start = min(usgs$measurement_date),
  measurement_end = max(usgs$measurement_date),
  minimum_head_ft_navd88 = min(usgs$groundwater_elevation_ft_navd88),
  maximum_head_ft_navd88 = max(usgs$groundwater_elevation_ft_navd88),
  coordinate_datum = paste(unique(usgs$coordinate_datum), collapse = ", ")
)
```

Download the [prepared attributed CSV](../downloads/usgs_hot_springs_stanley_shale_2017.csv)
and the [preparation script](../downloads/prepare_usgs_hot_springs.R). The DOI
release remains the authoritative source.

## Project the monitoring network

```{r prepare-public-points}
points <- vect(
  usgs,
  geom = c("longitude_nad83", "latitude_nad83"),
  crs = "EPSG:4269"
)
points <- project(points, "EPSG:26915")
points <- ps_make_points(
  points,
  value = "groundwater_elevation_ft_navd88",
  name_col = "site_id"
)

data.frame(
  points = nrow(points),
  projected_crs = crs(points, proj = TRUE),
  horizontal_units = "metres",
  head_units_and_datum = "feet above NAVD 88"
)
```

```{r public-network, fig.alt="Map of 36 USGS Stanley Shale monitoring wells east of Hot Springs, Arkansas, inside a buffered convex-hull analysis area in NAD83 UTM zone 15N."}
aoi <- buffer(convHull(points), 350)
plot(aoi, col = "#e8f3f6", border = "#176b87", lwd = 2,
     main = "USGS 2017 Stanley Shale monitoring network")
plot(points, add = TRUE, pch = 21, bg = "#f4b183", col = "#17252d")
```

The buffered convex hull is an example computational mask, not a mapped aquifer
boundary. Predictions near and beyond the outer wells require particular care.

## Interpolate, contour, and derive gradient direction

```{r public-surface-build}
surface <- ps_interpolate(
  points,
  methods = "TPS",
  grid_res = 75,
  mask = aoi,
  padding = 0
)$TPS
contours <- ps_contours(surface, interval = 20)
flow <- ps_flow_arrows(
  surface, res_factor = 12, scale = 10, min_gradient = 1e-5,
  endpoint_action = "shorten"
)
flow_tips <- ps_arrow_vertices(flow$arrows, "last")
tips_inside_map <- !is.na(extract(surface, flow_tips)[[2]])
flow$arrows <- flow$arrows[tips_inside_map, ]

data.frame(
  method = "TPS",
  grid_resolution_m = res(surface)[1],
  cells_with_values = global(surface, "notNA")[[1]],
  contour_features = nrow(contours),
  arrow_lines = nrow(flow$arrows)
)
```

```{r public-surface, fig.alt="TPS surface for 36 USGS Stanley Shale wells, shaded blue at lower modeled groundwater-level altitude and red at higher altitude, with twenty-foot contours and well locations."}
plot(surface, col = cols,
     main = "TPS example — groundwater-level altitude, ft above NAVD 88")
plot(contours, add = TRUE, col = "#17252d", lwd = 1)
plot(points, add = TRUE, pch = 21, bg = "white", cex = .65)
```

```{r public-flow, fig.alt="USGS Stanley Shale TPS surface with dark cartographic arrows pointing toward decreasing modeled groundwater-level altitude and twenty-foot contours."}
bases <- ps_arrow_vertices(flow$arrows, "first")
tips <- ps_arrow_vertices(flow$arrows, "last")
b <- crds(bases); t <- crds(tips)
plot(surface, col = cols,
     main = "Inferred hydraulic-gradient direction — software example")
plot(contours, add = TRUE, col = "#52646d", lwd = .8)
arrows(b[, 1], b[, 2], t[, 1], t[, 2], length = .055, col = "#111111")
plot(points, add = TRUE, pch = 21, bg = "white", cex = .55)
```

The official study used additional hydrogeologic interpretation and a different
surface-development workflow. These package results should not be used as a
replacement for that interpretation. The arrows are inferred gradient
directions from this interpolated surface; they are not velocities, travel
times, particle paths, or contaminant-transport trajectories.
