---
title: "Repeated monitoring events"
description: "Use ordinary R iteration around potentiomap for three synthetic monitoring rounds on one common grid."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repeated monitoring events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 9, fig.height = 5, dpi = 144, fig.align = "center")
library(potentiomap)
library(terra)
cols <- hcl.colors(64, "RdYlBu", rev = TRUE)
data("synthetic_wells", package = "potentiomap")
```

`potentiomap` operates on one observation set at a time. Base R splitting and
iteration can apply that released functionality to repeated events. This
article does not claim grouped processing as a built-in package feature.

## Create a reproducible long table

```{r long-table}
set.seed(20260529)
events <- as.Date(c("2026-01-15", "2026-04-15", "2026-07-15"))
long <- do.call(rbind, lapply(seq_along(events), function(i) {
  d <- synthetic_wells[c("well_id", "x", "y", "gw_elevation")]
  d$monitoring_date <- events[i]
  d$event_head <- d$gw_elevation - c(0, 0.7, 1.3)[i] +
    c(0, 0.00020, 0.00030)[i] * (d$x - mean(d$x))
  d
}))
long <- long[order(long$monitoring_date, long$well_id), ]
head(long[, c("monitoring_date", "well_id", "x", "y", "event_head")])
```

The changes are synthetic and intended only to demonstrate workflow structure.

## Apply a common AOI and grid

```{r interpolate-events}
aoi <- ps_sample_aoi()
template <- rast(crs = "EPSG:26916", ext = ext(aoi), resolution = 100)
by_event <- split(long, long$monitoring_date)

event_surfaces <- lapply(by_event, function(d) {
  points <- ps_make_points(
    d, "x", "y", "event_head", "well_id", "EPSG:26916"
  )
  ps_interpolate(points, methods = "TPS", template = template, mask = aoi)$TPS
})

do.call(rbind, lapply(names(event_surfaces), function(date) {
  r <- event_surfaces[[date]]
  data.frame(date = date, rows = nrow(r), columns = ncol(r),
             x_resolution_m = res(r)[1], finite_cells = global(r, "notNA")[[1]])
}))
```

```{r event-maps, fig.width=10, fig.height=4, fig.alt="Three synthetic monitoring-event TPS surfaces on the same grid and blue-low to red-high head scale, showing a designed decline and spatial shift through time."}
shared <- range(unlist(lapply(event_surfaces, minmax)), finite = TRUE)
par(mfrow = c(1, 3), mar = c(3, 3, 3, 1))
for (date in names(event_surfaces)) {
  plot(event_surfaces[[date]], col = cols, range = shared, main = date)
}
```

## Export event-specific products

```{r export-events}
event_root <- file.path(tempdir(), "potentiomap-repeated-events")
export_counts <- lapply(names(event_surfaces), function(date) {
  out <- ps_export_surfaces(
    setNames(list(event_surfaces[[date]]), "TPS"),
    out_dir = file.path(event_root, date),
    out_stub = paste0("event_", gsub("-", "", date)),
    contour_interval = 1,
    write_png = FALSE
  )
  data.frame(date = date, rasters = sum(file.exists(out$raster)),
             contour_datasets = sum(file.exists(out$contours)))
})
do.call(rbind, export_counts)
```

## Modeled-head difference

```{r difference-map, fig.alt="Difference raster of the final synthetic monitoring-event TPS surface minus the initial event, using blue for negative modeled-head differences and red for positive differences."}
modeled_difference <- event_surfaces[[3]] - event_surfaces[[1]]
limit <- max(abs(minmax(modeled_difference)))
plot(modeled_difference, col = hcl.colors(64, "Blue-Red 3", rev = TRUE),
     range = c(-limit, limit), main = "Final minus initial modeled head")
```

```{r difference-summary}
data.frame(
  comparison = paste(names(event_surfaces)[3], "minus", names(event_surfaces)[1]),
  minimum = minmax(modeled_difference)[1],
  mean = global(modeled_difference, "mean", na.rm = TRUE)[[1]],
  maximum = minmax(modeled_difference)[2]
)
```

An interpolated difference is a difference between two modeled surfaces. It is
not measured change at unsampled locations. Interpret it with measurement
timing, pumping conditions, network consistency, aquifer selection, uncertainty,
and any changed boundary conditions.
