---
title: "Troubleshooting and common errors"
description: "Observed potentiomap errors and practical checks for CRS, missing values, duplicates, kriging, grids, arrows, contours, and exports."
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting and common errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.bg = "white", fig.width = 8, fig.height = 5, dpi = 144)
library(potentiomap)
library(terra)
data("synthetic_wells", package = "potentiomap")
capture_error <- function(expr) tryCatch(expr, error = function(e) conditionMessage(e))
```

The messages below are captured from the released package; they are not
invented examples.

## Input and CRS errors

```{r observed-input-errors}
data.frame(
  situation = c("coordinate table without CRS", "missing groundwater-elevation column"),
  observed_message = c(
    capture_error(ps_make_points(synthetic_wells, "x", "y", "gw_elevation", "well_id")),
    capture_error(ps_make_points(synthetic_wells, "x", "y", "not_a_field", "well_id", "EPSG:26916"))
  )
)
```

Check `terra::is.lonlat()` and transform longitude/latitude observations to an
appropriate projected CRS before distance-based interpolation or support
classification. Assigning a projected CRS to longitude and latitude numbers is
not a transformation.

## Too few observations and unsupported methods

```{r observed-interpolation-errors}
points <- ps_make_points(
  synthetic_wells, "x", "y", "gw_elevation", "well_id", "EPSG:26916"
)
data.frame(
  situation = c("four observations", "unknown method"),
  observed_message = c(
    capture_error(ps_interpolate(points[1:4], methods = "IDW", grid_res = 100)),
    capture_error(ps_interpolate(points, methods = "UNKNOWN", grid_res = 100))
  )
)
```

Kriging can also fail or warn when the point count, spatial configuration, lag
classes, or empirical variogram do not support a stable fit. Preserve the
actual warning, inspect the variogram and fit, reconsider cutoff and width, and
do not silently substitute a preferred-looking surface.

## Missing values and duplicate coordinates

`ps_make_points()` drops rows whose coordinate or value cannot be converted to
a finite number. Compare row counts before and after standardization.

```{r missing-duplicate-check}
problem <- synthetic_wells
problem$gw_elevation[2] <- NA
problem <- rbind(problem, problem[1, ])
prepared <- ps_make_points(problem, "x", "y", "gw_elevation", "well_id", "EPSG:26916")
xy <- as.data.frame(crds(prepared))
data.frame(
  input_rows = nrow(problem),
  prepared_rows = nrow(prepared),
  duplicate_coordinate_rows = sum(duplicated(xy) | duplicated(xy, fromLast = TRUE))
)
```

Do not assume coincident observations are interchangeable. Review
whether duplicates represent repeat readings, nested wells, or data errors;
resolve them according to the monitoring design before interpolation.

## Smoothing and contour errors

```{r observed-output-errors}
surface <- ps_interpolate(points, methods = "IDW", grid_res = 100)$IDW
empty_surface <- surface
values(empty_surface) <- NA_real_
data.frame(
  situation = c("even smoothing window", "surface without finite values"),
  observed_message = c(
    capture_error(ps_smooth_surface(surface, window_size = 4)),
    capture_error(ps_contours(empty_surface, interval = 1))
  )
)
```

If contours are excessively dense or sparse, compare the interval with the
surface range and head precision. Explicit `levels` can be clearer than a very
small regular interval.

## AOI, grid size, and memory

Check that observations and the AOI overlap after transformation. An explicit
overlap check remains part of QA even when the interpolation method can return
finite predictions away from the monitoring network.

```{r grid-estimate}
aoi <- ps_sample_aoi()
width_m <- xmax(aoi) - xmin(aoi)
height_m <- ymax(aoi) - ymin(aoi)
resolutions <- c(200, 100, 25, 5)
data.frame(
  resolution_m = resolutions,
  approximate_cells = ceiling(width_m / resolutions) * ceiling(height_m / resolutions)
)
```

Cell count grows with the inverse square of cell size. Start with a resolution
supported by well spacing and intended map scale; test finer grids only when
there is a defensible need.

## Flat surfaces and arrow density

Flat or nearly flat gradients can be removed by `min_gradient`. Reduce it only
when small gradients are meaningful relative to measurement and interpolation
uncertainty. Increase `res_factor` to reduce arrow density; adjust `scale` for
cartographic length. Neither setting creates additional monitoring information.

## Export checks

- Test `file.access(output_directory, 2) == 0` before a long export.
- Review `overwrite` behavior and use a deliberate event-specific directory.
- Keep every `.shp`, `.shx`, `.dbf`, `.prj`, and `.cpg` component together.
- Read outputs back with `terra::rast()` or `terra::vect()` and compare CRS,
  dimensions, attribute names, and feature counts.
- If files become too large, revisit domain extent, raster resolution, number
  of methods, and whether every intermediate belongs in the deliverable.

## Quick diagnostic table

```{r troubleshooting-table, echo=FALSE}
knitr::kable(data.frame(
  symptom = c(
    "Missing or implausible map",
    "Kriging fit warning",
    "Too many or too few contours",
    "Cluttered or off-map arrows",
    "Unexpected export contents"
  ),
  first_check = c(
    "CRS, coordinate ranges, AOI overlap, and finite Z values",
    "Point count, spatial geometry, lag classes, variogram, and conditions",
    "Surface range, precision, interval, and requested-level manifest",
    "res_factor, scale, finite support, and endpoint validation",
    "Output directory, vector format, manifest, and read-back validation"
  )
))
```
